---
title: "biomass_across_salinity_treatment"
author: "Jared Umphress"
subtitle: "Created: September 27, 2026
           Last Updated: October 3, 2026 "
format: 
  pdf:
    colorlinks: true
toc: true

execute:
  warning: false
  message: false
---

comparing biomass based on salinity groups using the average ec measured



# Part 1: Set up tasks

Load in libraries
```{r}

library(tidyverse)
library(janitor)
library(here)

#plotting
library(cowplot)
library(ggthemes) # to use Paul Tol color palettes

# packages specifically for soil science
library(aqp) # algorithms for quantitative pedology
# library(soiltexture)

library(MuMIn) # model selection
library(ggeffects) # generating model predictions

library(ggplot2) # for ggplot
```

Load in data
```{r}

final_mass <- read_csv(here("data/stipa_salinity_experiment/final_mass_08_13.csv"))

final_ec <- read.csv(here("data/stipa_salinity_experiment/final_stipa_soil_ec.csv"))

```

# Part 2: Comparing the dry biomasses

### Convert stipa ec values to dS/m and account for water dilution
```{r}

source(here("code/mS_per_cm_to_dS_per_m.R"))

final_ec <- final_ec %>%
  clean_names() %>%
  mutate(
    ec_ds_per_m = uS_per_cm_to_dS_per_m(ec_u_s_m),
    ec_converted = (ec_u_s_m/1000)*5,
    year = year(date_measured),
    treatment = paste(as.character(salinity_trt), water_trt)
  ) %>%
  arrange(desc(salinity_trt))
  

```


Get average ec for each salinity group. Disregarding dry vs wet. So should be four groups: 0.8, 4, 7.5, 11
```{r}

final_ec <- final_ec %>%
  group_by(salinity_trt) %>%
  mutate(pooled_salinity_trt = mean(ec_converted))

# 11 intended = 11.81 actual
# 7.5 intended = 9.98 actual
# 4 intended = 5.58 actual
# 0.8 intended = 2.98 actual

```



## Add the actual treatment group (actual ec + water treatment) for the plants based on their ID's
```{r}

final_mass <- final_mass %>%
    mutate(pooled_salinity_trt = case_when(
    plant_id >= 1 & plant_id <= 20 ~ "2.98",
    plant_id >= 21 & plant_id <= 40 ~ "5.58",
    plant_id >= 41 & plant_id <= 60 ~ "9.98",   # this is actually 7.5 intended
    plant_id >= 61 & plant_id <= 80 ~ "11.81",
 )) %>%
  mutate(total_mass = root_dry_mass_g + live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  mutate(all_shoot_mass = live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  mutate(root_to_shoot = root_dry_mass_g / (live_shoot_dry_mass_g + dead_shoot_dry_mass_g))

```


## Get mean total_biomass per salinity group
```{r}

mass_summary <- final_mass %>%
  group_by(pooled_salinity_trt) %>%
  summarise(
    mean_total_mass = mean(total_mass, na.rm = TRUE),
    sd_total_mass = sd(total_mass, na.rm = TRUE),
    
    mean_live_shoot_mass = mean(live_shoot_dry_mass_g, na.rm = TRUE),
    sd_live_shoot_mass = sd(live_shoot_dry_mass_g, na.rm = TRUE),
    
    mean_all_shoot_mass = mean(all_shoot_mass, na.rm = TRUE),
    sd_all_shoot_mass = sd(all_shoot_mass, na.rm = TRUE),
    
    mean_root_mass = mean(root_dry_mass_g, na.rm = TRUE),
    sd_root_mass = sd(root_dry_mass_g, na.rm = TRUE),
    
    mean_root_to_shoot = mean(root_to_shoot, na.rm = TRUE),
    sd_root_to_shoot = sd(root_to_shoot, na.rm = TRUE),
    
    n_sampled = n(),
  ) %>%
  arrange(pooled_salinity_trt)

```


## 2.1: Compare total mass across salinity treatments
Visualize the total masses of the plants across salinity treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = pooled_salinity_trt,
                     y = total_mass,
                     color = pooled_salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = pooled_salinity_trt, 
                                # mean computer time is the y value
                                y = mean_total_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_total_mass + sd_total_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_total_mass - sd_total_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total mass (g)",
       title = "Total Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```

- anova/wilcoxon assumptions
- run anova




## 2.2: Compare live shoot mass across salinity treatments
Visualize live shoot across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = pooled_salinity_trt,
                     y = live_shoot_dry_mass_g,
                     color = pooled_salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = pooled_salinity_trt, 
                                # mean computer time is the y value
                                y = mean_live_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_live_shoot_mass + sd_live_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_live_shoot_mass - sd_live_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total live shoot mass (g)",
       title = "Total Live Shoot Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## 2.3: Compare root mass across salinity treatments
Visualize root mass across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = pooled_salinity_trt,
                     y = root_dry_mass_g,
                     color = pooled_salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = pooled_salinity_trt, 
                                # mean computer time is the y value
                                y = mean_root_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_mass + sd_root_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_mass - sd_root_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total root mass (g)",
       title = "Total Root Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## 2.4: Compare all shoot mass across salinity treatments
Visualize all shoot mass across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = pooled_salinity_trt,
                     y = all_shoot_mass,
                     color = pooled_salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = pooled_salinity_trt, 
                                # mean computer time is the y value
                                y = mean_all_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_all_shoot_mass + sd_all_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_all_shoot_mass - sd_all_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total all shoot mass (g)",
       title = "Total All Shoot Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## 2.5: Compare root to shoot ratio across salinity treatments
Visualize root to shoot ratio across salinity treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = pooled_salinity_trt,
                     y = root_to_shoot,
                     color = pooled_salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = pooled_salinity_trt, 
                                # mean computer time is the y value
                                y = mean_root_to_shoot,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_to_shoot + sd_root_to_shoot, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_to_shoot - sd_root_to_shoot)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Root to Shoot Ratio",
       title = "Root to Shoot Ratio across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


# Conclusions

The salinity group of 9.98 seems to have the highest total biomass, live shoot, root mass. All root to shoot ratios are comparble. The results seem to suggest that 3 dS and 6 dS are too low for our local genotype. 11.81 dS is too high. But 10 dS on average is justttt right. 