---
title: "biomass_vs_water_treatment"
author: "Jared Umphress"
subtitle: "Created: October 1, 2026
           Last Updated: October 3, 2026 "
format: 
  pdf:
    colorlinks: true
toc: true

execute:
  warning: false
  message: false
---

# Part 1: Set up tasks

Load in libraries
```{r}

library(tidyverse)
library(janitor)
library(here)

#plotting
library(cowplot)
library(ggthemes) # to use Paul Tol color palettes

# packages specifically for soil science
library(aqp) # algorithms for quantitative pedology
# library(soiltexture)

library(MuMIn) # model selection
library(ggeffects) # generating model predictions

library(ggplot2) # for ggplot
library(rstatix) # for cohens_d test
```


Load in data
```{r}

final_mass <- read_csv(here("data/stipa_salinity_experiment/final_mass_08_13.csv"))

```


### Add the actual treatment group (actual ec + water treatment) for the plants based on their ID's
```{r}

final_mass <- final_mass %>%
  mutate(salinity_trt = case_when(
    plant_id >= 1 & plant_id <= 10 ~ "3.513",
     plant_id >= 11 & plant_id <= 20 ~ "2.426",
     plant_id >= 21 & plant_id <= 30 ~ "6.490",
     plant_id >= 31 & plant_id <= 40 ~ "4.666",
     plant_id >= 41 & plant_id <= 50 ~ "10.201",   # this is actually 7.5 intended
     plant_id >= 51 & plant_id <= 60 ~ "9.755",
     plant_id >= 61 & plant_id <= 70 ~ "10.039",
     plant_id >= 71 & plant_id <= 80 ~ "13.590"
  )) %>%
  mutate(water_trt = case_when(
     plant_id >= 1 & plant_id <= 10 ~ "wet",
     plant_id >= 11 & plant_id <= 20 ~ "dry",
     plant_id >= 21 & plant_id <= 30 ~ "wet",
     plant_id >= 31 & plant_id <= 40 ~ "dry",
     plant_id >= 41 & plant_id <= 50 ~ "wet",
     plant_id >= 51 & plant_id <= 60 ~ "dry",
     plant_id >= 61 & plant_id <= 70 ~ "wet",
     plant_id >= 71 & plant_id <= 80 ~ "dry"
  )) %>%
  mutate(total_mass = root_dry_mass_g + live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  mutate(all_shoot_mass = live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  mutate(root_to_shoot = root_dry_mass_g / (live_shoot_dry_mass_g + dead_shoot_dry_mass_g))

```

# Part 2: Compare masses based on water treatment

### Get mean biomasses per water treatment group (wet/dry)
```{r}

water_mass_summary <- final_mass %>%
  group_by(water_trt) %>%
  summarise(
    mean_total_mass = mean(total_mass, na.rm = TRUE),
    sd_total_mass = sd(total_mass, na.rm = TRUE),
    n_total_mass = n(),
    
    mean_live_shoot_mass = mean(live_shoot_dry_mass_g, na.rm = TRUE),
    sd_live_shoot_mass = sd(live_shoot_dry_mass_g, na.rm = TRUE),
    n_live_shoot_mass = n(),
    
    mean_all_shoot_mass = mean(all_shoot_mass, na.rm = TRUE),
    sd_all_shoot_mass = sd(all_shoot_mass, na.rm = TRUE),
    n_all_shoot_mass = n(),
    
    mean_root_mass = mean(root_dry_mass_g, na.rm = TRUE),
    sd_root_mass = sd(root_dry_mass_g, na.rm = TRUE),
    n_root_mass = n(),
    
    mean_root_to_shoot = mean(root_to_shoot, na.rm = TRUE),
    sd_root_to_shoot = sd(root_to_shoot, na.rm = TRUE),
    n_root_to_shoot = n(),
    
    n_sampled = n(),
  ) 

```

## 2.1: Total Mass
Visualize the total masses of the plants across WATER treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = total_mass,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_total_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_total_mass + sd_total_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_total_mass - sd_total_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total mass (g)",
       title = "Total Mass across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


```{r}
#| label: histogram

# Rice rule formula 2n^(1/3) where n = number sampled
2*(34)^(1/3) 
# 6.47

ggplot(data = final_mass, # starting data frame
       aes(x = total_mass)) + # x-axis (no y-axis for histogram)
  geom_histogram(bins = 6) + # number of bins from Rice Rule
  facet_wrap(~water_trt) # creating two panels to show watersheds separately
```


```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_mass, # starting data frame
       aes(sample = total_mass)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "blue") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```

```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  total_mass ~ water_trt, # formula: response variable ~ grouping variable
  data = final_mass # data: final_ec
)
```
We determined that group variances were not equal (F-test, F(29, 29) = 0.43, p = 0.02).


```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  total_mass ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_mass
)

```
We did not find a significant difference in total mass between wet
treated (n = 34, mean = 2.26 grams) and dry (n = 34, mean = 1.95 grams)
treated (Welch’s t-test, t(57.45) = 3.8, p = 0.28, 𝛼 = 0.05).

```{r}
#| label: cohen's d

cohens_d(
  total_mass ~ water_trt,
  data = final_mass
  #pooled_sd = FALSE
)

```
Cohen's d was 0.267. So water treatment had a small effect size. 

## 2.2: Live shoot
Visualize live shoot across water treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = live_shoot_dry_mass_g,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_live_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_live_shoot_mass + sd_live_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_live_shoot_mass - sd_live_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total live shoot mass (g)",
       title = "Total Live Shoot Mass across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


```{r}
#| label: histogram

# Rice rule formula 2n^(1/3) where n = number sampled
2*(34)^(1/3) 
# 6.47

ggplot(data = final_mass, # starting data frame
       aes(x = live_shoot_dry_mass_g)) + # x-axis (no y-axis for histogram)
  geom_histogram(bins = 6) + # number of bins from Rice Rule
  facet_wrap(~water_trt) # creating two panels to show watersheds separately
```
Both histograms are skewed left.


```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_mass, # starting data frame
       aes(sample = live_shoot_dry_mass_g)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "tomato2") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```
Data is not that normally distributed. 40 sample size so can relax normality?

```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  live_shoot_dry_mass_g ~ water_trt, # formula: response variable ~ grouping variable
  data = final_mass # data: final_ec
)
```
We determined that group variances were not equal (F-test, F(30, 30) = 0.71, p = 0.35).


```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  live_shoot_dry_mass_g ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_mass
)

```
p - value is 0.951. No significant difference in total mass based on water treatment


## 2.3: Root mass
Visualize root mass
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = root_dry_mass_g,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_root_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_mass + sd_root_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_mass - sd_root_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total root mass (g)",
       title = "Total Root Mass across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


```{r}
#| label: histogram

# Rice rule formula 2n^(1/3) where n = number sampled
2*(34)^(1/3) 
# 6.47

ggplot(data = final_mass, # starting data frame
       aes(x = root_dry_mass_g)) + # x-axis (no y-axis for histogram)
  geom_histogram(bins = 6) + # number of bins from Rice Rule
  facet_wrap(~water_trt) # creating two panels to show watersheds separately
```
Pretty normal and the total sample is 34, so you can apply central limit theorem. 


```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_mass, # starting data frame
       aes(sample = root_dry_mass_g)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "orchid") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```
Data follows qq line pretty well will some skew at the ends. Can conclude that data is normally distributed

```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  root_dry_mass_g ~ water_trt, # formula: response variable ~ grouping variable
  data = final_mass # data: final_ec
)
```
We determined that group variances were not equal (F-test, F(33, 33) = 0.68, p = 0.28).


```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  root_dry_mass_g ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_mass
)

```
p - value is 0.5171. The difference in root dry mass between wet and dry treatment groups is not statistically significant.


## 2.4: All shoot mass (live + dead)
Visualize all shoot mass across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = all_shoot_mass,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_all_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_all_shoot_mass + sd_all_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_all_shoot_mass - sd_all_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total all shoot mass (g)",
       title = "Total All Shoot Mass across water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


```{r}
#| label: histogram

# Rice rule formula 2n^(1/3) where n = number sampled
2*(34)^(1/3) 
# 6.47

ggplot(data = final_mass, # starting data frame
       aes(x = all_shoot_mass)) + # x-axis (no y-axis for histogram)
  geom_histogram(bins = 6) + # number of bins from Rice Rule
  facet_wrap(~water_trt) # creating two panels to show watersheds separately
```
Pretty normal


```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_mass, # starting data frame
       aes(sample = all_shoot_mass)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "skyblue") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```
Data follows qq line moderately well. 


```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  all_shoot_mass ~ water_trt, # formula: response variable ~ grouping variable
  data = final_mass # data: final_ec
)
```
We determined that group variances were not equal (F-test, F(29, 29) = 0.32, p < 0.05).


```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  all_shoot_mass ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_mass
)

```
p - value = 0.1902. Difference in all shoot mass is not statistically significant.


## 2.5: Root to shoot
Visualize root to shoot ratio across salinity treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = root_to_shoot,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_root_to_shoot,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_to_shoot + sd_root_to_shoot, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_to_shoot - sd_root_to_shoot)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Root to Shoot Ratio",
       title = "Root to Shoot Ratio across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


```{r}
#| label: histogram

# Rice rule formula 2n^(1/3) where n = number sampled
2*(34)^(1/3) 
# 6.47

ggplot(data = final_mass, # starting data frame
       aes(x = root_to_shoot)) + # x-axis (no y-axis for histogram)
  geom_histogram(bins = 6) + # number of bins from Rice Rule
  facet_wrap(~water_trt) # creating two panels to show watersheds separately
```
Wet has an outlier on the right. 


```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_mass, # starting data frame
       aes(sample = root_to_shoot)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "chocolate") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```
Data follows qq line moderately well. For wet group there is big skew towards the upper end. 


```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  root_to_shoot ~ water_trt, # formula: response variable ~ grouping variable
  data = final_mass # data: final_ec
)
```
We determined that group variances were not equal (F-test, F(29, 29) = 0.69, p = 0.31).


```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  root_to_shoot ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_mass
)

```
p - value = 0.8817. The difference in root to shoot ratio between wet vs dry groups was not statistically significant. 


# Conclusions
Write this out in fuller sentences later. With proper statistical wording.

When comparing the stipa pulchra groups in wet (n = ), vs dry (n = ), we did not find a statistically significant difference in total mass, live mass, root mass, nor root to shoot ratio. 

Interpretation: Our data and analysis suggest that stipa pulchra with "wet feet" vs regular watering did not have an impact on the successful growth of stipa pulchra.





