Created by Jared Umphress

binomial logistic regression for survival across diff salinities
visually compare total biomass, root to shoot ratio using jitter for the four salinity treatments

statistically compare total biomass, root to shoot ratio using anova or non-parametric version

---
title: "stipa_expirement_analysis"
author: "Jared Umphress"
subtitle: "Created: September 27, 2026
           Last Updated: October 1, 2026 "
format: 
  pdf:
    colorlinks: true
toc: true

execute:
  warning: false
  message: false
---

# Part 1: Set up tasks

Load in libraries
```{r}

library(tidyverse)
library(janitor)
library(here)

#plotting
library(cowplot)
library(ggthemes) # to use Paul Tol color palettes

# packages specifically for soil science
library(aqp) # algorithms for quantitative pedology
# library(soiltexture)

library(MuMIn) # model selection
library(ggeffects) # generating model predictions

library(ggplot2) # for ggplot
```

Load in data
```{r}

final_ec <- read.csv(here("data/stipa_salinity_experiment/final_stipa_soil_ec.csv"))

```
The stipa ec values from 07/10/26 were only one plant per group. Not enough to get a good idea of the salnity per group. It serves as a check in on the expirement. It was the salniity when we applied the saline treatment 2 of the 3 days. After harvesting we started to water 3 days a week. 

# Part 2: EC Values

### Convert stipa ec values to dS/m and account for water dilution
```{r}

source(here("code/mS_per_cm_to_dS_per_m.R"))

final_ec <- final_ec %>%
  clean_names() %>%
  mutate(
    ec_ds_per_m = uS_per_cm_to_dS_per_m(ec_u_s_m),
    ec_converted = (ec_u_s_m/1000)*5,
    year = year(date_measured),
    treatment = paste(as.character(salinity_trt), water_trt)
  ) %>%
  arrange(desc(salinity_trt))
  

```


### Get average ec value per treatment group
```{r}

mean_ec_values <- final_ec %>%
  group_by(treatment) %>%
  summarise(
    mean_ec = mean(ec_converted),
    sd_ec = sd(ec_converted),
    n_sampled = n()
  )

```

### Create column for average ec value per treatment group
```{r}

final_ec <- final_ec %>%
  mutate(
    actual_treatment = case_match(
      treatment,
      "11 wet" ~ "10.039 dS wet",
      "11 dry" ~ "13.59 dS dry",
      "7.5 wet" ~ "10.201 dS wet (7.5)",
      "7.5 dry" ~ "9.755 dS dry",
      "4 wet" ~ "6.490 dS wet",
      "4 dry" ~ "4.666 dS dry",
      "0.8 dry" ~ "2.426 dS dry",
      "0.8 wet" ~ "3.513 dS wet"
    ))

```
This is the actual ec value per treatment group. This included the water treatment and intended salinity. 

### Create plot of actual ec values vs target
```{r}

target_ec <- final_ec %>%
  select(salinity_trt, ec_converted, water_trt)

ggplot(
  data = target_ec,
  aes(x = salinity_trt,
                y = ec_converted,
      color = water_trt)) +
    geom_point() +
    geom_abline(slope = 1) +
  labs(x = "Intended Salinity (dS)",
       y = "Actual Salinity (dS)")

# 36 of the 40 measured soils salinities surpased the intended salinity 

```

## Part 2.2: Compare wet vs dry treatments

### are wet treatments on average more saline than the dry treatments?
```{r}

wet_v_dry <- final_ec %>%
  group_by(water_trt) %>%
  summarise(mean_ec = mean(ec_converted),
            sd_ec = sd(ec_converted),
            n_sampled = n())

# mean_dry = 7.60
# mean_wet = 7.56
# the mean ec_values are pretty similar
# can run a t.test to compare

```

### Visualize mean and sd of each water treament 
```{r}

# base layer is ggplot
# data is final_ec
ggplot(data = final_ec,
       mapping = aes(x = water_trt,
                     y = ec_converted,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = wet_v_dry,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_ec,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_ec + sd_ec, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_ec - sd_ec))
  # manually setting colors for the treatment groups
  scale_color_manual(values = c("dry" = "yellow3",
                                "wet" = 'skyblue2')) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Salinity (dS/m)",
       title = "Title"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```

```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_ec, # starting data frame
       aes(sample = ec_converted)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "blue") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```

```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  ec_converted ~ water_trt, # formula: response variable ~ grouping variable
  data = final_ec # data: final_ec
)
```
#### F statistic is 2.3021. Variances are not equal.

```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  ec_converted ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_ec
)
```
#### We determined that the mean soil ec values between the wet water treatments and dry water treatments were not different. 


# Part 3: Comparing the dry biomasses

## Load in data
```{r}

final_mass <- read_csv(here("data/stipa_salinity_experiment/final_mass_08_13.csv"))

```


## Add the actual treatment group (actual ec + water treatment) for the plants based on their ID's
```{r}

final_mass <- final_mass %>%
  mutate(salinity_trt = case_when(
    plant_id >= 1 & plant_id <= 10 ~ "3.513",
     plant_id >= 11 & plant_id <= 20 ~ "2.426",
     plant_id >= 21 & plant_id <= 30 ~ "6.490",
     plant_id >= 31 & plant_id <= 40 ~ "4.666",
     plant_id >= 41 & plant_id <= 50 ~ "10.201",   # this is actually 7.5 intended
     plant_id >= 51 & plant_id <= 60 ~ "9.755",
     plant_id >= 61 & plant_id <= 70 ~ "10.039",
     plant_id >= 71 & plant_id <= 80 ~ "13.590"
  )) %>%
  mutate(water_trt = case_when(
     plant_id >= 1 & plant_id <= 10 ~ "wet",
     plant_id >= 11 & plant_id <= 20 ~ "dry",
     plant_id >= 21 & plant_id <= 30 ~ "wet",
     plant_id >= 31 & plant_id <= 40 ~ "dry",
     plant_id >= 41 & plant_id <= 50 ~ "wet",
     plant_id >= 51 & plant_id <= 60 ~ "dry",
     plant_id >= 61 & plant_id <= 70 ~ "wet",
     plant_id >= 71 & plant_id <= 80 ~ "dry"
  )) %>%
  mutate(total_mass = root_dry_mass_g + live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  mutate(all_shoot_mass = live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  mutate(root_to_shoot = root_dry_mass_g / (live_shoot_dry_mass_g + dead_shoot_dry_mass_g))

```

# Part 3.1: Compare biomasses based on salinity treatments 
## Get mean total_biomass per salinity group
```{r}

mass_summary <- final_mass %>%
  group_by(salinity_trt) %>%
  summarise(
    mean_total_mass = mean(total_mass, na.rm = TRUE),
    sd_total_mass = sd(total_mass, na.rm = TRUE),
    
    mean_live_shoot_mass = mean(live_shoot_dry_mass_g, na.rm = TRUE),
    sd_live_shoot_mass = sd(live_shoot_dry_mass_g, na.rm = TRUE),
    
    mean_all_shoot_mass = mean(all_shoot_mass, na.rm = TRUE),
    sd_all_shoot_mass = sd(all_shoot_mass, na.rm = TRUE),
    
    mean_root_mass = mean(root_dry_mass_g, na.rm = TRUE),
    sd_root_mass = sd(root_dry_mass_g, na.rm = TRUE),
    
    mean_root_to_shoot = mean(root_to_shoot, na.rm = TRUE),
    sd_root_to_shoot = sd(root_to_shoot, na.rm = TRUE),
    
    n_sampled = n(),
  ) %>%
  arrange(salinity_trt)

```


## Visualize the total masses of the plants across salinity treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = salinity_trt,
                     y = total_mass,
                     color = salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = salinity_trt, 
                                # mean computer time is the y value
                                y = mean_total_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_total_mass + sd_total_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_total_mass - sd_total_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total mass (g)",
       title = "Total Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize live shoot across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = salinity_trt,
                     y = live_shoot_dry_mass_g,
                     color = salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = salinity_trt, 
                                # mean computer time is the y value
                                y = mean_live_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_live_shoot_mass + sd_live_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_live_shoot_mass - sd_live_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total live shoot mass (g)",
       title = "Total Live Shoot Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize root mass across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = salinity_trt,
                     y = root_dry_mass_g,
                     color = salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = salinity_trt, 
                                # mean computer time is the y value
                                y = mean_root_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_mass + sd_root_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_mass - sd_root_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total root mass (g)",
       title = "Total Root Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize all shoot mass across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = salinity_trt,
                     y = all_shoot_mass,
                     color = salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = salinity_trt, 
                                # mean computer time is the y value
                                y = mean_all_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_all_shoot_mass + sd_all_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_all_shoot_mass - sd_all_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Total all shoot mass (g)",
       title = "Total All Shoot Mass across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize root to shoot ratio across salinity treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = salinity_trt,
                     y = root_to_shoot,
                     color = salinity_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = salinity_trt, 
                                # mean computer time is the y value
                                y = mean_root_to_shoot,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_to_shoot + sd_root_to_shoot, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_to_shoot - sd_root_to_shoot)) +
  # changing axis labels and title
  labs(x = "Salinity Treatment",
       y = "Root to Shoot Ratio",
       title = "Root to Shoot Ratio across Salinity Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


# Part 3.2: Compare masses based on water treatment

## Get mean biomasses per salinity group
```{r}

water_mass_summary <- final_mass %>%
  group_by(water_trt) %>%
  summarise(
    mean_total_mass = mean(total_mass, na.rm = TRUE),
    sd_total_mass = sd(total_mass, na.rm = TRUE),
    
    mean_live_shoot_mass = mean(live_shoot_dry_mass_g, na.rm = TRUE),
    sd_live_shoot_mass = sd(live_shoot_dry_mass_g, na.rm = TRUE),
    
    mean_all_shoot_mass = mean(all_shoot_mass, na.rm = TRUE),
    sd_all_shoot_mass = sd(all_shoot_mass, na.rm = TRUE),
    
    mean_root_mass = mean(root_dry_mass_g, na.rm = TRUE),
    sd_root_mass = sd(root_dry_mass_g, na.rm = TRUE),
    
    mean_root_to_shoot = mean(root_to_shoot, na.rm = TRUE),
    sd_root_to_shoot = sd(root_to_shoot, na.rm = TRUE),
    
    n_sampled = n(),
  ) 

```

## Visualize the total masses of the plants across WATER treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = total_mass,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_total_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_total_mass + sd_total_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_total_mass - sd_total_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total mass (g)",
       title = "Total Mass across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```

```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = final_mass, # starting data frame
       aes(sample = total_mass)) + # y-axis for QQ plot (no x-axis)
  # first layer: QQ reference line 
  geom_qq_line(color = "blue") + # showing this in blue so it's easier to see
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~water_trt) # show watersheds separately

```

```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  total_mass ~ water_trt, # formula: response variable ~ grouping variable
  data = final_mass # data: final_ec
)
```
#### F statistic is 2.3021. Variances are not equal.

```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  total_mass ~ water_trt, 
  # argument for equal/unequal variances (variances should be equal)
  var.equal = FALSE, 
  # data: final_ec
  data = final_mass
)

```
# low statistical difference in mean total mass between water treatments (p = 0.2847)



## Visualize live shoot across water treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = live_shoot_dry_mass_g,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_live_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_live_shoot_mass + sd_live_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_live_shoot_mass - sd_live_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total live shoot mass (g)",
       title = "Total Live Shoot Mass across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize root mass
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = root_dry_mass_g,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_root_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_mass + sd_root_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_mass - sd_root_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total root mass (g)",
       title = "Total Root Mass across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize all shoot mass across salinity treatments
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = all_shoot_mass,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_all_shoot_mass,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_all_shoot_mass + sd_all_shoot_mass, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_all_shoot_mass - sd_all_shoot_mass)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Total all shoot mass (g)",
       title = "Total All Shoot Mass across water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```


## Visualize root to shoot ratio across salinity treatments 
```{r}

# base layer is ggplot
# data is final_mass
ggplot(data = final_mass,
       mapping = aes(x = water_trt,
                     y = root_to_shoot,
                     color = water_trt
                     )) +
  # second layer is a jitter plot
  geom_jitter(
              # width is set to 0.15 to tighten the horizontal jitter
              width = 0.17,  
               # height is set to 0 so that there is no vertical jitter
              height = 0,  
              # open circle point shape for jitter points
              shape = 21,   
               # increasing transparency so that it is easier to see the mean
              alpha = 0.8  
             ) +  
   geom_pointrange(data = water_mass_summary,  
                  # points displayed based on weekday vs weekend grouping
                  mapping = aes(x = water_trt, 
                                # mean computer time is the y value
                                y = mean_root_to_shoot,   
                                # upper bound is the mean plus 
                                #  a standard deviation
                                ymax = mean_root_to_shoot + sd_root_to_shoot, 
                                # lower bound is the mean minus 
                                #  a standard deviation
                                ymin = mean_root_to_shoot - sd_root_to_shoot)) +
  # changing axis labels and title
  labs(x = "Water Treatment",
       y = "Root to Shoot Ratio",
       title = "Root to Shoot Ratio across Water Treatments"
      ) +
  # changing theme from default theme to theme_classic
  # theme_classic creates a white panel with no grid lines
  theme_classic() +
  # removing the legend 
  theme(legend.position = "none")

```



# Compare mean total mass of groups with ANOVA or equivalent



# number of individuals per group over time




# binomial logistic regression for survival



