---
title: "survival_binomial_regression"
format: html
---

- in final mass dataframe use mutate and if else statement to create survival column 
- divide plants into pooled_salinity_trt groups: 2.98, 5.58, 9.98, 11.31
- check assumptions of binomial logistic regression
- run a binomial logistic regression 

# Part 1: Set up tasks

Load in libraries
```{r}

library(tidyverse)
library(janitor)
library(here)

#plotting
library(cowplot)
library(ggthemes) # to use Paul Tol color palettes

# packages specifically for soil science
library(aqp) # algorithms for quantitative pedology
# library(soiltexture)

library(MuMIn) # model selection
library(ggeffects) # generating model predictions

library(ggplot2) # for ggplot
library(DHARMa) # for logistic regression diagnostics
library(flextable)

```

Load in data
```{r}

final_mass <- read_csv(here("data/stipa_salinity_experiment/final_mass_08_13.csv"))

final_ec <- read.csv(here("data/stipa_salinity_experiment/final_stipa_soil_ec.csv"))

```

# Part 2: Comparing the dry biomasses

### Convert stipa ec values to dS/m and account for water dilution
```{r}

source(here("code/mS_per_cm_to_dS_per_m.R"))

final_ec <- final_ec %>%
  clean_names() %>%
  mutate(
    ec_ds_per_m = uS_per_cm_to_dS_per_m(ec_u_s_m),
    ec_converted = (ec_u_s_m/1000)*5,
    year = year(date_measured),
    treatment = paste(as.character(salinity_trt), water_trt)
  ) %>%
  arrange(desc(salinity_trt))
  

```


Get average ec for each salinity group. Disregarding dry vs wet. So should be four groups: 0.8, 4, 7.5, 11
```{r}

final_ec <- final_ec %>%
  group_by(salinity_trt) %>%
  mutate(pooled_salinity_trt = mean(ec_converted))

# 11 intended = 11.81 actual
# 7.5 intended = 9.98 actual
# 4 intended = 5.58 actual
# 0.8 intended = 2.98 actual

```



## Add the actual treatment group (actual ec + water treatment) for the plants based on their ID's
```{r}

final_mass <- final_mass %>%
    mutate(pooled_salinity_trt = case_when(
    plant_id >= 1 & plant_id <= 20 ~ "2.98",
    plant_id >= 21 & plant_id <= 40 ~ "5.58",
    plant_id >= 41 & plant_id <= 60 ~ "9.98",   # this is actually 7.5 intended
    plant_id >= 61 & plant_id <= 80 ~ "11.81",
 )) %>%
  # creting total biomass column
  mutate(total_mass = root_dry_mass_g + live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  # creating column representing total mass of the live plus dead shoot dry mass
  mutate(all_shoot_mass = live_shoot_dry_mass_g + dead_shoot_dry_mass_g) %>%
  # creating column of root to shoot ratio
  mutate(root_to_shoot = root_dry_mass_g / (live_shoot_dry_mass_g + dead_shoot_dry_mass_g)) %>%
  # creating binary column of survival (1) vs death (0)
  mutate(survival = case_when(
    live_shoot_dry_mass_g == 0 ~ "0",
    live_shoot_dry_mass_g > 0 ~ "1"
  ))

```

- Homoscedastic errors

Assumptions: 
- Independent observations = yes
- Response/dependent variable is distributed following an exponential
family distribution (e.g. normal/Gaussian, binomial, Poisson)
- Linear relationship between transformed expected response in terms
of link function and explanatory variables
Departures from what we’ve seen before:
- Homogeneity of variance doesn’t need to be satisfied (and it usually
isn’t, given model structure)
- Errors need to be independent but not necessarily normally
distributed

```{r}

# since I made the survival column using case_match, which writes in values as strings
# I need to make the values of 0 and 1 as.numeric
final_mass$survival <- as.numeric(final_mass$survival)

# since I made the pooled_salinity_trt column using case_match, which writes in values as strings
# I need to make the values of 0 and 1 as.numeric
final_mass$pooled_salinity_trt <- as.numeric(final_mass$pooled_salinity_trt)


# running general linearilized model
survival_model_1 <- glm(
  # survival as response to the different pooled_salinity_trt groups
  survival ~ pooled_salinity_trt,
  # using final_mass data
  data = final_mass,
  # running the binomial model
  family = "binomial"
)



```


look at diagnostics
```{r}

# plot QQ plot 
# also runs the KS test, dispersion test, outlier test
# n.s. result indicates that the results are not significant and not likely to affect 
plot(
  simulateResiduals(survival_model_1)
)

```

What each test means and the results
- KS test (Kolmogorov-Smirnov): do
your residuals follow a uniform
distribution? Not significant
- Dispersion test: are your residuals
more dispersed than expected from a
uniform distribution? Not significant
- Outlier test: are there outliers in
your data set that could be
influencing model coefficients or
predictions? Not significant

```{r}

summary(survival_model_1)

# what does plogis do?
plogis(0)
plogis(10)

```


Look at the model predictions
```{r}


ggpredict(survival_model_1,
          terms = "pooled_salinity_trt [5]")

# If the salinity is 5 dS then the probability that the plant will survive is 0.88

```


Visualize model predictions
```{r}

mod_preds <- ggpredict(survival_model_1,
                       terms = "pooled_salinity_trt [0:15 by = 0.5]")

ggplot(data = final_mass,
       aes(x = pooled_salinity_trt,
           y = survival)) + 
  geom_point(size = 3,
             alpha = 0.5) +
  geom_ribbon(data = mod_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high,
              alpha = 0.4)) +
  geom_line(data = mod_preds,
            aes(x = x, 
                y = predicted)) +
  scale_y_continuous(limits = c(0,1),
                     breaks = c(0,1))


```


```{r}

# creating flextable to show intercept estimate, slope estimate from saliniity effects. standard error, and z value
as_flextable(survival_model_1)

```



