---
title: "Summer 2026 Grassland Soil Core Figures"
author: "Jared Umphress"
subtitle: "Created: August 16, 2026
           Last Updated: September 24, 2026 "
format: 
  pdf:
    colorlinks: true
toc: true

execute:
  warning: false
  message: false
---

# Part 1: Set up tasks

Load in libraries
```{r}

library(tidyverse)
library(janitor)
library(here)

#plotting
library(cowplot)
library(ggthemes) # to use Paul Tol color palettes

# packages specifically for soil science
library(aqp) # algorithms for quantitative pedology
# library(soiltexture)

library(MuMIn) # model selection
library(ggeffects) # generating model predictions

library(ggplot2) # for ggplot
```

Load in data
```{r}

soil_data <- read.csv(here("data/soil_samples/2026_summer_grassland_soil_cores.csv"))

```

# Part 2: Convert EC measurements to dS/m 

This function also accounts for dilution

```{r}

source(here("code/mS_per_cm_to_dS_per_m.R"))

#clean data, get depth midpoints, and estimate EC in dS/m
soil_data <- soil_data %>% 
  clean_names() %>% 
  mutate(
         depth_midpoint = -(depth_top + depth_bottom)/2,
         ec_ds_per_m = uS_per_cm_to_dS_per_m(ec_u_s_cm),
         ec_converted = (ec_u_s_cm/1000)*5,
         year = year(sample_date),
          # use lapply to apply the conversion function to the vector
         ssc = lapply(texture_code, texcl_to_ssc),
         #subset ssc to the corresponding columns
         sand = sapply(ssc, `[[`, "sand"),
         silt = sapply(ssc, `[[`, "silt"),
         clay = sapply(ssc, `[[`, "clay"),
         year = year(sample_date)
  ) %>%
  select(-ssc) %>%  # Remove temporary column
  #move quantitative columns next to texture
  relocate(sand:clay, .after = texture_code)

```

Test uS/cm to dS/cm conversion
```{r}

# high value of 2440
uS_per_cm_to_dS_per_m(2440)
# 16.451 dS/cm

# low value of 235 uS/cm
uS_per_cm_to_dS_per_m(235)
# 3.005 dS/cm

```

## 2.2: Get mean EC, soil moisture, and sand/clay content values for depth midpoints 

```{r}

#calculate mean values for both plots combined
means_2026_summer <- soil_data %>% 
  group_by(depth_midpoint) %>% 
  summarize(mean_ec = mean(ec_ds_per_m, na.rm = TRUE),
            mean_moisture = mean(percent_soil_moisture),
            mean_sand = mean(sand),
            mean_clay = mean(clay),
            n = n()) %>% 
  ungroup()

```


# Part 3: Plot EC, soil moisture, clay/sand percentage by depth based on transect 

## 3.1 EC figures
```{r}

fig_2026_summer_ec <- ggplot(data = soil_data, aes(x = depth_midpoint, y = ec_ds_per_m, color = transect)) +
  geom_point() +
  geom_path() +
  geom_path(data = means_2026_summer, aes(x = depth_midpoint, y = mean_ec), color = "black", linewidth = 2) +
  coord_flip() +
  theme_cowplot() +
  scale_x_continuous(limits = c(-90,0), breaks = seq(-90, 0, by =15)) +
  scale_y_continuous(limits = c(0,NA), position = "right") +
  labs(y = "EC (dS/m)", x = "Depth (cm)", title = "Mesa Grassland Cores, Summer 2026") +
  scale_color_ptol()

fig_2026_summer_ec

#save to file
ggsave(filename = here("figures/soil_cores/2026/summer/Summer_2026_Mesa_Grassland_EC.pdf"),
       fig_2026_summer_ec,
       width = 6,
       height = 4,
       units = "in",
       bg = "white")


```

## 3.2 Soil Moisture figures
```{r}

#  moisture plot ---- 
fig_2026_summer_moisture <- ggplot(data = soil_data, aes(x = depth_midpoint, y = percent_soil_moisture, color = transect)) +
  geom_point() +
  geom_path() +
  # plot thick black line for mean values
  geom_path(data = means_2026_summer, aes(x = depth_midpoint, y = mean_moisture), color = "black", linewidth = 2) +
  coord_flip() +
  theme_cowplot() +
  scale_x_continuous(limits = c(-90,0), breaks = seq(-90, 0, by =15)) +
  scale_y_continuous(limits = c(0,NA), position = "right") +
  labs(y = "% moisture", x = "Depth (cm)", title = "Mesa Grassland Cores, Summer 2026") +
  scale_color_ptol()

fig_2026_summer_moisture

ggsave(filename = here("figures/soil_cores/2026/summer/Summer_2026_Mesa_Grassland_moisture.pdf"),
       fig_2026_summer_moisture,
       width = 6,
       height = 4,
       units = "in",
       bg = "white")

```

## 3.3: Plot Average Sand% and Average Clay % by Depth
```{r}

# Texture  ----

##   Sand  ----
fig_2026_summer_sand <- ggplot(data = soil_data, aes(x = depth_midpoint, y = sand)) +
  geom_point(aes(color = transect, group = transect)) +
  #geom_jitter() +
  geom_path(aes(color = transect, group = transect)) +
  geom_path(data = means_2026_summer, aes(x = depth_midpoint, y = mean_sand), color = "black", linewidth = 2) +
  theme_cowplot() +
  coord_flip() +
  #ylab("Sample depth (cm)") +
  #xlab("% Sand (estimated)") +
  scale_x_continuous(limits = c(-90,0), breaks = seq(-90, 0, by =15)) +
  scale_y_continuous(limits = c(0,100), position = "right") +
  scale_color_ptol() +
  labs(title = "Mesa Grasslands, Summer 2026", y = "Sand (%)", x = "Depth (cm)")

fig_2026_summer_sand

## Clay  ----
fig_2026_summer_clay <- ggplot(data = soil_data, aes(x = depth_midpoint, y = clay)) +
  geom_point(aes(color = transect, group = transect)) +
  #geom_jitter() +
  geom_path(aes(color = transect, group = transect)) +
  geom_path(data = means_2026_summer, aes(x = depth_midpoint, y = mean_clay), color = "black", linewidth = 2) +
  theme_cowplot() +
  coord_flip() +
  #ylab("Sample depth (cm)") +
  #xlab("% Sand (estimated)") +
  scale_x_continuous(limits = c(-90,0), breaks = seq(-90, 0, by =15)) +
  scale_y_continuous(limits = c(0,100), position = "right") +
  scale_color_ptol() +
  labs(title = "Mesa Grasslands, Summer 2026", y = "Clay (%)", x = "Depth (cm)")

fig_2026_summer_clay

## combined plot ----
fig_slope_texture <- plot_grid(fig_2026_summer_sand, fig_2026_summer_clay,
                            nrow = 1)
fig_slope_texture

#save to file
ggsave(filename = "figures/soil_cores/2026/summer/Summer_2026_Mesa_Grassland_texture.pdf", 
       plot = fig_slope_texture,
       width = 8,
       height = 5,
       units = "in")

```

Notes:
* GL10 has pretty homogenous soil texture of sandy clay. Its soil moisture is relatively homogenous throughout the various depths. EC does not skew until last depth 

## 3.4 Code to look at one transect at a time
```{r}

# using the soil_data, create smaller data frame called one_core_data 
one_core_data <- soil_data %>%
  # filter to include only the transects of interest
  filter(transect == "GL-2")


one_core_2026_summer <- ggplot(data = one_core_data, 
                                   aes(x = depth_midpoint, 
                                       y = percent_soil_moisture, # ec_ds_per_m or percent_soil_moisture
                                       color = transect)) +
  geom_point() +
  geom_path() +
  geom_path(data = means_2026_summer, aes(x = depth_midpoint, y = mean_ec), color = "black", linewidth = 1.5) +
  coord_flip() +
  theme_cowplot() +
  scale_x_continuous(limits = c(-90,0), breaks = seq(-90, 0, by =15)) +
  scale_y_continuous(limits = c(0,NA), position = "right") +
  labs(y = "EC (dS/m)", x = "Depth (cm)", title = "One Mesa Grassland Core, Summer 2026") +
  scale_color_ptol()

one_core_2026_summer

```


# Part 5: Get mean EC, soil moisture, and sand/clay content values for each transect
```{r}

soil_data_summary <- soil_data %>%
  group_by(transect) %>%
  summarise(
    mean_ec = mean(ec_u_s_cm),
    mean_soil_moisture = mean(percent_soil_moisture),
    mean_clay_content = mean(clay),
    mean_sand_content = mean(sand)
  )

soil_data_summary

```

## 5.2: Create bar charts displaying means
```{r}

### for soil electrical conductivity (salinity) ###
mean_ec_graph <- ggplot(
  data = soil_data_summary,
  aes(
      x = transect, 
      y = mean_ec,
      fill = transect
)) +
  geom_col() 
 
mean_ec_graph


### for clay content ###
mean_clay_graph <- ggplot(
  data = soil_data_summary,
  aes(
    x = transect,
    y = mean_clay_content, 
    fill = transect
  )
) +
  geom_col()

mean_clay_graph


### for soil sand content ###
mean_sand_graph <- ggplot(
  data = soil_data_summary,
   aes(
    x = transect,
    y = mean_sand_content, 
    fill = transect
  )
) +
  geom_col()

mean_sand_graph

# GL-9 stands out as a transect with the highest clay content and lowest sand content

### for soil percent moisture ###
mean_soil_moisture_graph <- ggplot(
  data = soil_data_summary,
  aes(
    x = transect,
    y = mean_soil_moisture,
    fill = transect
  )
) +
  geom_col()

mean_soil_moisture_graph


```

For each transect, I will plot where I took the core using the coordinates. Using the average soil conductivities I can divide it by some factor to create the size of the soil core point to visually represent the differences in soil EC at each transect


# Part 6: Linear model trials 

## 6.1 Check normality of ec_ds_per_m, clay content, and soil moisture
```{r}

# ggplot as base layer
ggplot(
  # using kelp data
  data = soil_data,
  # plotting temperature 
  mapping = aes(sample = ec_ds_per_m)
) +
  # qq line is first layer
  geom_qq_line() +
  # qq points are second layer
  # changing color of the points
  geom_qq(color = 'yellow3') +
  # setting theme as minimal
  theme_cowplot()

# ggplot as base layer
ggplot(
  # using kelp data
  data = soil_data,
  # plotting temperature 
  mapping = aes(sample = clay)
) +
  # qq line is first layer
  geom_qq_line() +
  # qq points are second layer
  # changing color of the points
  geom_qq(color = 'brown') +
  # setting theme as minimal
  theme_cowplot()

# ggplot as base layer
ggplot(
  # using kelp data
  data = soil_data,
  # plotting temperature 
  mapping = aes(sample = percent_soil_moisture)
) +
  # qq line is first layer
  geom_qq_line() +
  # qq points are second layer
  # changing color of the points
  geom_qq(color = 'blue1') +
  # setting theme as minimal
  theme_cowplot()

```

## 6.2 Create linear models
```{r}

# model 1 #
model_1 <- lm(
  ec_ds_per_m ~ clay, # formula: change in soil EC as a function of clay percentage
  data = soil_data      # using soil_data 
)

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
  # if you don't use par the plots will not be displayed at the same time
plot(model_1)    # will display diagnostic plots 


# model 2 #
model_2 <- lm(
  ec_ds_per_m ~ percent_soil_moisture, # formula: change in soil EC as a function of percent soil moisture 
  data = soil_data      # using soil_data '
)

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
  # if you don't use par the plots will not be displayed at the same time
plot(model_2)    # will display diagnostic plots 


# model 3 #
model_3 <- lm(
  percent_soil_moisture ~ clay, # formula: change in soil_moisture as a function of clay content
  data = soil_data      # using soil_data '
)

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
  # if you don't use par the plots will not be displayed at the same time
plot(model_3)    # will display diagnostic plots 


# model 4 
model_4 <- lm(
  ec_ds_per_m ~ percent_soil_moisture + clay, # formula: change in soil ec as a function of soil moisture and clay content without an interaction
  data = soil_data      # using soil_data '
)

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
  # if you don't use par the plots will not be displayed at the same time
plot(model_4)    # will display diagnostic plots 


# model 5 
model_5 <- lm(
  ec_ds_per_m ~ percent_soil_moisture * clay, # formula: change in soil ec as a function of soil moisture interacting with clay content
  data = soil_data      # using soil_data '
)

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
  # if you don't use par the plots will not be displayed at the same time
plot(model_5)    # will display diagnostic plots 

# model 6
model_6 <- lm(
  ec_ds_per_m ~ percent_soil_moisture * clay + sand,
  data = soil_data
)

par(mfrow = c(2, 2))   # displays plots in a 2x2 grid
  # if you don't use par the plots will not be displayed at the same time
plot(model_6)    # will display diagnostic plots 

```

## 6.3 Run AICc test
```{r}

AICc(
  model_1,
  model_2,
  model_3,
  model_4,
  model_5,
  model_6
) |>
  arrange(AICc) # arrange AICc values by descending order 

# The model with the lowest AICc is the strongest relationship
# model_5 has the lowest AICc

```

Get a summary of Model 5
```{r}

summary(model_5)

```

## 6.4 Get model predictions 
```{r}
#| label: model_5 predictions

# creating a new object called model_2_preds
model_5_preds <- ggpredict(
  model_5,      # model object
  terms = "percent_soil_moisture"   # predictor (in quotation marks)
)

# display the predictions
model_5_preds
# this object will be used to create the regression line and its ribbon

```

## Plotting linear model 
```{r}
#| label: soil_moisture vs ec_u_s_cm -model-final-figure

# base layer: ggplot
# using clean data frame
ggplot(data = soil_data,
       aes(x = percent_soil_moisture,
           y = ec_u_s_cm)) +
  # first layer: points representing abalones
  geom_point(size = 2,
             stroke = 1,
             fill = "firebrick4",
             shape = 21) +
  # second layer: ribbon representing confidence interval
  # using predictions data frame
  geom_ribbon(data = model_5_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.1) +
  # third layer: line representing model predictions
  # using predictions data frame
 geom_line(data = model_5_preds,
            aes(x = x,
                y = predicted)) +
  # axis labels
  labs(x = "Percent Soil Moisture", 
       y = "Soil EC") +
  # theme
  theme_minimal()
```

The line of best fit and ribbnon are bellow all the actual points. Maybe the relationship is logarithmic?

# Part 7: Is The Relationship Logarithmic?
This part is expirementational 

```{r}

model_6 <- lm(log(ec_u_s_cm) ~ log(percent_soil_moisture), data = soil_data )

summary(model_6)

```

```{r}
#| label: model_6 predictions

# creating a new object called model_2_preds
model_6_preds <- ggpredict(
  model_6,      # model object
  terms = "percent_soil_moisture"   # predictor (in quotation marks)
)

# display the predictions
model_6_preds

```

```{r}
#| label: soil_moisture vs ec_u_s_cm -model-final-figure logarithmic

# base layer: ggplot
# using clean data frame
ggplot(data = soil_data,
       aes(x = percent_soil_moisture,
           y = ec_u_s_cm)) +
  # first layer: points representing abalones
  geom_point(size = 2,
             stroke = 1,
             fill = "skyblue2",
             shape = 21) +
  # second layer: ribbon representing confidence interval
  # using predictions data frame
  geom_ribbon(data = model_6_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.1) +
  # third layer: line representing model predictions
  # using predictions data frame
 geom_line(data = model_6_preds,
            aes(x = x,
                y = predicted)) +
  # axis labels
  labs(x = "Percent Soil Moisture", 
       y = "Soil EC") +
  # theme
  theme_minimal()
```

